---
title: "LAB5"
author: "Murat Oztok"
date: "4/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#LOAD THE DATA and IMPORT
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(tidyverse)
library(lubridate)
library(readxl)
library(ggrepel)
states <- map_data("state")
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv" , stringsAsFactors = FALSE)
names_of_the_states <- states %>% group_by(region) %>% summarize(long=mean(long), lat=mean(lat))
glc <- readxl::read_xlsx("~/Desktop/DS202/WEEK12/LAB5/FRPP_GLC_-_United_StATESaPRIL122021_(2).xlsx")
```
<br>
<br>

#Cleaning and Preprocessing
```{r}
str(glc)
glc$STATE = as.numeric(glc$`State Code`)
glc_states <- glc %>% group_by(STATE) %>% summarize(state_name=first(`State Name`))
acc_with_state_names <- acc %>% left_join(glc_states, by=c('STATE'))

```


#1 Are there some days of the week where more accidents happen than the others ?
```{r}
table(acc$DAY_WEEK)
# As we can see no 7 has the highest value which corresponds to Saturday, then Sunday and Friday follows. We can generalize and say more accidents happen in weekends than weekdays.
```
<br>
<br>

#2 Create a data frame containing the persons who are fatally hurt in the accidents.
```{r}
fatally_hurt <- per %>% filter(INJ_SEV == 4)
```
<br>
<br>


#3 create a data frame containing the most dangerous vehicle make in each state.
```{r}
new_dat <- acc_with_state_names %>% inner_join(per, by=c('ST_CASE', 'STATE')) %>% select(STATE, state_name, MAKE)
most_dangerous <- new_dat %>%
  filter(!is.na(MAKE)) %>%
  group_by(STATE, state_name, MAKE) %>%
  summarize(number_accidents=n()) %>%
  filter(number_accidents==max(number_accidents))
most_dangerous
```
<br>
<br>


#4 Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map.
```{r}
#Lets make sure our region names are uppercase in our state df.
states$region = toupper(states$region)
avg_state_locations <- states %>% group_by(region) %>% summarize(avg_long=mean(long), avg_lat=mean(lat))
dangerous_makes_with_locations <- most_dangerous %>% mutate(region=state_name) %>% inner_join(avg_state_locations, by=c('region'))
ggplot(states, aes(x=long, y=lat)) + geom_polygon(aes(group=group)) +
  geom_text_repel(data=dangerous_makes_with_locations, aes(x=avg_long, y=avg_lat, label=MAKE), color='blue')
#The definition of most dangerous vehicles can be automobile makers who had involved in the highest number of fatal accidents.
#In this map, we can the MAKE code for the most dangerous vehicle by each states
#Looking at the map, we see many 12 which corresponds to Ford, 20s  which corresponds to Chevrolet

```
<br>
<br>


#5 Join the accident and person table.
```{r}
accidents_people <- acc %>% inner_join(per)
```
<br>
<br>


#6 Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.
```{r}
accidents_people %>% group_by(SEX) %>%
  summarize(number_accidents=n()) %>%
  ggplot(aes(x=SEX, weight=number_accidents)) + geom_bar() +
  xlab('Sex Attribute Code') + ylab('No. of Accidents')
acc %>% group_by(HOUR) %>%
  filter(HOUR != 99) %>% #Outliers
  summarize(number_accidents=n()) %>%
  ggplot(aes(x=HOUR, weight=number_accidents)) + geom_bar() +
  xlab('Hours in Day') + ylab('No. of Accidents')
acc %>% group_by(DAY_WEEK) %>%
  summarize(number_accidents=n()) %>%
  mutate(weekday=c('Sunday', 'Monday', 'Tuesday', 'Wedneday', 'Thursday', 'Friday', 'Saturday')) %>%
  select(-DAY_WEEK) %>%
  subset(select=c(2,1)) %>%
  ggplot(aes(x=weekday, weight=number_accidents)) + geom_bar() +
  xlab('Days') + ylab('No. of Accidents')
#Summaries above show that males are likely to be in an accident that females.
#The most common hour is 6pm for fatal accidents.
#Saturday is the most common day where fatal accidents occur.

```
<br>
<br>


#7 Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.
```{r}
counties <- map_data('county')
deaths_by_county <- accidents_people  %>% 
  filter(INJ_SEV==4) %>% 
  group_by(STATE, COUNTY) %>%
  summarize(total_deaths=n())
glc_counties <- glc %>%
  mutate(STATE=as.numeric(`State Code`), COUNTY=as.numeric(`County Code`), region=tolower(`State Name`), subregion=tolower(`County Name`)) %>%
  group_by(region, subregion, STATE, COUNTY) %>%
  summarize(region=first(region), subregion=first(subregion), STATE=first(STATE), COUNTY=first(COUNTY))
deaths_by_county_with_names <- deaths_by_county %>% left_join(glc_counties, by=c('STATE', 'COUNTY'))
dat <- counties %>% left_join(deaths_by_county_with_names, by=c('region', 'subregion'))
ggplot(dat, aes(x=long, y=lat, fill=total_deaths)) + geom_polygon(aes(group=group))


#Here is a map of  total deaths by county. Grayed areas had no records in the dataset implying they had no recorded fatality accidents.


```
<br>
<br>


#8 Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}
summer <- acc_with_state_names %>% filter(MONTH %in% c(6,7,8)) #Summer
winter <- acc_with_state_names %>% filter(MONTH %in% c(12, 1, 2)) #Winter
nrow(summer)
nrow(winter)
summer1 <- summer %>% group_by(state_name) %>% summarize(num_summer_accidents=n())
winter1 <- winter %>% group_by(state_name) %>% summarize(num_winter_accidents=n())
dat <- inner_join(summer1, winter1)
dat %>% filter(num_winter_accidents > num_summer_accidents)
dat

#More fatal accidents during summer, generalization can be made by saying summers are more dangerous than winters.
#Looking into the data more, we can see the data suggests that this depend on the state. 

```
